---
title: "Project 2"
author: "Brian Kreis"
date: "October 11, 2015"
output: html_document
---
##Crime Stats

Loading of libraries and initial cleanup work
```{r}
library(tidyr)
library(dplyr)
library(ggplot2)

crime <- read.csv("https://raw.githubusercontent.com/bkreis84/Project-2/master/table_12_crime_trends_by_population_group_2012-2013.csv", header = TRUE, skip = 3, check.names = FALSE, stringsAsFactors = FALSE)

colnames(crime)[2] <- "Year"

crime <- crime[1:42,1:14]
tbl_df(crime)

```


We do not care about the percent change in our analysis so we can remove this using a logical vector. 

```{r}
crimex <- crime[c(TRUE, TRUE, FALSE),]
```

We now replace blanks with NA values so that we can then use the fill function to fill in missing values using the value above.

```{r}
crimex[crimex==""] <- NA

crimex <- crimex %>% 
  fill(`Population group`) 
```

After noticing some errors (likely spaces on the end of character strings), we change our variable names. 

We then select the specific variables we want using the "select" function. We then use the arrange function to sort our data in a way that we can once again use the fill function to fill in missing values with values from above. 

```{r}
colnames(crimex)[3] <- "Violent crime"
colnames(crimex)[8] <- "Property crime"
colnames(crimex)[14] <- "2013 estimated population"

crimex <- select(crimex, Year,`Population group`, `Violent crime`, `Property crime`, `2013 estimated population`)

crimex <- arrange(crimex, `Population group`, desc(Year)) 

crimex <- crimex %>% 
  fill(`2013 estimated population`)
```

I found a function to trim data using regular expressions. I hope to learn more about functions in the coming weeks. 
http://r.789695.n4.nabble.com/Remove-space-from-string-td4292042.html

Once again remove data not pertinant to our analysis.

```{r}
trim <- function(x) gsub(",|^[[:space:]]+|[[:space:]]+$", "", x) 
crimex <- data.frame(sapply(crimex, trim), check.names = FALSE)
#realizing that I need to dive into functions more

crimex <- crimex[-19:-28,]
crimex <- crimex[-1:-6,]
```

Change our character values to numeric now that they have been trimmed. This step took me quite awhile because every time I tried to convert the values to numeric, they were filled in with NA values. This is why I needed the trim function above (there were characters and spaces preventing the conversion)

```{r}
crimex$`Violent crime` <- as.numeric(as.character(crimex$`Violent crime`))
crimex$`Property crime` <- as.numeric(as.character(crimex$`Property crime`))
crimex$`Population group` <- as.character(crimex$`Population group`)
crimex$`2013 estimated population` <- as.numeric(as.character(crimex$`2013 estimated population`))

vcrime <- select(crimex, -`Property crime`)
pcrime <- select(crimex, -`Violent crime`)
```

I spread the data in order to get the year as a variable (I later determined that this did not really tidy the data). I also created new variables which showed the rate of violent crime per 100,000 population. 

```{r}           
vcrime <- spread(vcrime, Year, `Violent crime`)

vcrime <- mutate(vcrime, `2012 Violence Rate` = vcrime$`2012` / (vcrime$`2013 estimated population` / 100000))
vcrime <- mutate(vcrime, `2013 Violence Rate` = vcrime$`2013` / (vcrime$`2013 estimated population` / 100000))
```

```{r}
# Realized that the crimex set is actually the tidy one, so used that for the graph.
crimex <- mutate(crimex, `Violence Rate` = crimex$`Violent crime` / (crimex$`2013 estimated population` / 100000))
crimex$`Violence Rate` <- as.integer(crimex$`Violence Rate`)
tbl_df(crimex)

ggplot(crimex, aes(x=Year, y=`Violence Rate`, color=`Population group`, group=`Population group`)) +
  geom_point(size=2) +
  geom_line(size=2)
```

##MLB Stats

This one is relatively straightforward.We begin with pulling in the data and removing unnecessary columns

```{r}
wins <- read.csv("https://raw.githubusercontent.com/bkreis84/Project-2/master/wins.csv")
wins <- wins[,-5]
wins <- wins[,-2]
```

We then use the gather function to tidy up or data.

```{r}
wins <- gather(wins, "Team", "Wins", 2:31)
```

This graph is relatively useless! Just for practice

```{r}
ggplot(wins, aes(x=Year, y=Wins, color=Team, group=Team)) +
  geom_point(size=2) +
  geom_line(size=2)
```


##Giants & Jets
Again we start with bringing in the data and doing some intial cleanup

```{r}
giants <- read.csv("https://raw.githubusercontent.com/bkreis84/Project-2/master/Jets%20Giants.csv")
tbl_df(giants)


giants[3:5, "Team"] <- "Jets"
giants[8:10, "Team"] <- "Giants"

c("Team", "Stats", "2007", "2008", "2009", "2010", "2011", "2012", "2013", "2014") -> colnames(giants)

giants <- giants[-1,]
giants <- giants[-5,]
```

We use the gather and spread functions to tidy the data and mutate to create a new variable.

```{r}
giants <- gather(giants, "Year", "n", 3:10)

giants <- spread(giants, "Stats", n)
tbl_df(giants)

giants <- giants %>% 
  mutate(`Total Yards` = `Total Passing Yards` + `Total Rushing Yards`)
```  
  
We then use various dplyr functions to obtain information for our analysis
  
```{r}  
giants1 <-  giants %>% 
  group_by(Team, Year) %>% 
  summarise (`Passing Ratio` = `Total Passing Yards` / `Total Yards`,
             `Rushing Ratio` = `Total Rushing Yards` / `Total Yards`,
             `TD/1stDown` = Touchdowns / `Total First Downs`)
tbl_df(giants1)

giants2 <- giants %>% 
  group_by(Team) %>% 
  summarise(`Max TD` = max(Touchdowns), `Min TD` = min(Touchdowns), 
            `Avg Rushing Yards` = mean(`Total Rushing Yards`),
            `Avg Passing Yards` = mean(`Total Passing Yards`),
            `Avg Total Yards` = mean(`Total Yards`))
tbl_df(giants2)

giants3 <- select(giants, Year, Team, contains("Yards"))
tbl_df(giants3)
```  

Finally we create a couple charts to show the Total yards of each team over the years and the passing ratio. The information here conclusively proves that The Giants > Jets.

```{r}  
ggplot(giants3, aes(x = Year, y = `Total Yards`, color = Team, group = Team)) +
  geom_point(size = 8) + 
  geom_line(size = 2) 

ggplot(giants1, aes(x = Year, y = `Passing Ratio`, color = Team, group = Team)) +
  geom_point(size = 8) + 
  geom_line(size = 2) 
            
```
