---
title: "Week 7"
author: "Brian Kreis"
date: "October 16, 2015"
output: html_document
---

First, we load our required packages and data.

```{r}
library(jsonlite)
library(XML)
library(RCurl)

xbooks <- "https://raw.githubusercontent.com/bkreis84/Kreis-Week-7/master/books.xml"
jbooks <- "https://raw.githubusercontent.com/bkreis84/Kreis-Week-7/master/books.json"
hbooks<- "https://raw.githubusercontent.com/bkreis84/Kreis-Week-7/master/books.html"
```

I used the getUrl function in order to make the HTTPS request. The xmlToDataFrame function converts the xml file into an R data frame. The result is sloppy, in that it puts all of the authors into a single column and doesn't recognize the "subchildren". I could have input each author as a child, in which case all 3 tables would have been identical. Just for the sake of seeing what it would look like, I went this way.

```{r}
xbooks <- getURL(xbooks)
xframe <- xmlToDataFrame(xbooks)
xframe
```

JSON was very straightforward in that a single function was able to create the R table.

```{r}
jframe <- fromJSON(jbooks)
jframe
```

Again we used the getURL function, followed by a relatively simple fucntion to read the HTML information and convert it into a data frame. 

```{r}
hbooks <- getURL(hbooks)
hframe <- data.frame(readHTMLTable(hbooks))
hframe
```

The html and JSON data fraas I mentioned, I easily could have left each author as a child and had all of the tables match.
